---
title: "A Priori Power Analysis: Rogers et al. (2017) Study 1 Close Replication"
author: "Ke Fang"
date: "2026-06-04"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(pwr)

GLOBAL_SEED   <- 42
set.seed(GLOBAL_SEED)

alpha         <- 0.05
target_powers <- c(0.80, 0.90, 0.95)
```

# Overview

A close replication of Rogers, Zeckhauser, Gino, Norton & Schweitzer (2017),
Study 1 ("Artful Paltering", *JPSP* 112(3), 456–473). We power the three tests
the authors actually used:

1. **Choice rate** — proportion choosing to mislead, palter vs. lie-by-commission.
   Reported as a χ²(1) on the 2×2 table, which is algebraically the two-proportion
   *z* test, so we power the proportion difference directly. Original: 71% (palter)
   vs. 55% (lie), χ²(1) = 14.58, *p* < .001.
2. **Ethicality index** — independent-samples *t* test. Original: *M* = 4.05
   (*SD* 1.36) vs. *M* = 2.12 (*SD* 1.42); Cohen's *d* = 1.39.
3. **Mediation** — condition → ethicality → choice with a *binary* outcome
   (logistic *b*-path), indirect effect via a Preacher & Hayes bootstrap. Original
   logit model: *b*-path (ethicality → choice) = 0.455 (*SE* .077); *c′*-path
   (direct condition → choice) = 0.16 (*SE* .269, *n.s.*).

Tests 1 and 2 use closed-form power (`pwr` / base R). Test 3 uses Monte Carlo
simulation, because the outcome is binary and the indirect effect is non-normal:
we scan *N* with the joint-significance test (a fast, well-validated proxy for
bootstrap power; MacKinnon et al. 2002; Fritz & MacKinnon 2007) and confirm at the
chosen *N* with the actual nonparametric bootstrap of *a·b*. Two-sided, α = .05
throughout, matching the original.

# Effect-size inputs

Taken directly from the published paper.

```{r inputs}
## Test 1: proportion choosing the MISLEADING option, by condition
p_palter <- 0.71      # Palter condition
p_lie    <- 0.55      # Lie-by-Commission condition

## Test 2: ethicality index (1-7 scale), by condition
m_palter <- 4.05; sd_palter <- 1.36
m_lie    <- 2.12; sd_lie    <- 1.42

## Test 3: mediation path estimates (condition coded Palter = 1, Lie = 0)
a_path      <- m_palter - m_lie   # condition -> ethicality (OLS slope) = 1.93
b_path      <- 0.455              # ethicality -> choice (logit), reported
cprime_path <- 0.16               # direct condition -> choice (logit), n.s.
```

# Test 1 — Choice-rate difference

The χ²(1) on the 2×2 is the two-proportion *z* test, so we power the proportion
comparison. Three convergent methods, plus a cross-check against the reported χ².

```{r test1}
## Method A: Cohen's h (the standard ES for two independent proportions)
h <- ES.h(p_palter, p_lie)                       # 2*asin(sqrt p1) - 2*asin(sqrt p2)
n_arcsine <- sapply(target_powers, function(pw)
  ceiling(pwr.2p.test(h = h, sig.level = alpha, power = pw,
                      alternative = "two.sided")$n))

## Method B: base R, exact binomial-variance normal approximation
n_baseR <- sapply(target_powers, function(pw)
  ceiling(power.prop.test(p1 = p_palter, p2 = p_lie, sig.level = alpha,
                          power = pw, alternative = "two.sided")$n))

## Method C: chi-square effect size w on the 2x2 (what they reported).
## Build the 2x2 JOINT probability table assuming equal allocation (0.5/0.5).
joint <- matrix(c(0.5 * p_palter, 0.5 * (1 - p_palter),    # Palter row
                  0.5 * p_lie,    0.5 * (1 - p_lie)),       # Lie row
                nrow = 2, byrow = TRUE,
                dimnames = list(c("Palter", "Lie"),
                                c("Mislead", "Truth")))
w <- ES.w2(joint)                                # Cohen's w for a 2-way table
N_chisq_total <- sapply(target_powers, function(pw)
  ceiling(pwr.chisq.test(w = w, df = 1, sig.level = alpha, power = pw)$N))

cat(sprintf("Cohen's h = %.3f ;  Cohen's w (from proportions) = %.3f\n", h, w))

test1 <- data.frame(
  power               = target_powers,
  n_per_group_arcsine = n_arcsine,
  n_per_group_baseR   = n_baseR,
  N_total_baseR       = 2 * n_baseR,
  N_total_chisq       = N_chisq_total          # should match 2 * n_per_group
)
test1
```

As a cross-check, the effect size implied by the *reported* χ²(1) = 14.58 at
*N* = 550 (rather than recomputed from the rounded proportions):

```{r test1-chisq-crosscheck}
w_from_chisq <- sqrt(14.58 / 550)
cat(sprintf("Cohen's w from reported chi2/N = %.3f\n", w_from_chisq))
data.frame(
  power   = target_powers,
  N_total = sapply(target_powers, function(pw)
    ceiling(pwr.chisq.test(w = w_from_chisq, df = 1, sig.level = alpha, power = pw)$N))
)
```

The two routes to *w* agree closely (≈ 0.166 from the proportions, ≈ 0.163 from
χ²/N); they imply ~286 vs. ~296 total at 80% power. The discrepancy is just
rounding — the published 0.71/0.55 in a balanced *N* = 550 table imply χ² ≈ 15.10,
not 14.58. We treat the proportion-based figure as primary for transparency.

# Test 2 — Ethicality index (between-subjects *t* test)

```{r test2}
## Recompute Cohen's d from the reported Ms/SDs (pooled SD, equal n assumed)
sd_pooled <- sqrt((sd_palter^2 + sd_lie^2) / 2)
d         <- (m_palter - m_lie) / sd_pooled
cat(sprintf("Pooled SD = %.3f ;  Cohen's d = %.3f\n", sd_pooled, d))

n_t <- sapply(target_powers, function(pw)
  ceiling(pwr.t.test(d = d, sig.level = alpha, power = pw,
                     type = "two.sample", alternative = "two.sided")$n))

test2 <- data.frame(power = target_powers, n_per_group = n_t, N_total = 2 * n_t)
test2
```

The ethicality effect is enormous (*d* ≈ 1.4), so this test needs only a handful
per cell and never binds the design.

# Test 3 — Mediation (binary outcome, simulation)

Generative model (condition *X*: Lie = 0, Palter = 1):

$$M \mid X \sim \text{Normal}(b_{0M} + a\,X,\ \sigma_M), \qquad
  Y \mid X,M \sim \text{Bernoulli}\!\big(\text{logit}^{-1}(b_{0Y} + b\,M + c'X)\big).$$

We take *a*, σ_M, *b*, and *c′* from the paper and solve *b₀Y* so the model
reproduces the observed overall mislead rate (~0.63). Because logistic
coefficients are non-collapsible, the model-implied per-condition rates are *close
to* but not exactly 0.71 / 0.55 — expected and harmless for power, which depends
on a realistic *a*-path, *b*-path, and *Y* base rate. The script prints the
implied rates as a sanity check.

```{r test3-model}
pars <- list(a = a_path, b = b_path, cprime = cprime_path,
             b0M = m_lie, sdM = sd_pooled, b0Y = NA_real_)

## Solve b0Y to match the overall mislead rate, integrating M out with a fixed
## (deterministic) standard-normal grid so uniroot sees a smooth function.
solve_b0Y <- function(target_overall, p, n_quad = 1e5) {
  z <- qnorm(ppoints(n_quad))               # deterministic, symmetric grid
  x <- rep(0:1, each = n_quad / 2)
  m <- p$b0M + p$a * x + p$sdM * z
  f <- function(b0Y) mean(plogis(b0Y + p$b * m + p$cprime * x)) - target_overall
  uniroot(f, interval = c(-50, 50))$root
}
overall_rate <- (p_palter + p_lie) / 2       # = 0.63
pars$b0Y <- solve_b0Y(overall_rate, pars)
cat(sprintf("Solved intercept b0Y = %.3f\n", pars$b0Y))

## Sanity check: model-implied P(mislead) by condition
implied_by_condition <- function(p, n = 2e5) {
  x  <- rep(0:1, each = n / 2)
  m  <- rnorm(n, p$b0M + p$a * x, p$sdM)
  pr <- plogis(p$b0Y + p$b * m + p$cprime * x)
  setNames(tapply(pr, x, mean), c("Lie(X=0)", "Palter(X=1)"))
}
cat("Model-implied mislead rates (target 0.55 / 0.71):\n")
print(round(implied_by_condition(pars), 3))
```

The simulation and estimation helpers: one dataset, its analysis (*a*-path by OLS,
*b*-path by logistic regression, indirect = *a·b*), and the two power routines
(joint-significance and the nonparametric bootstrap of *a·b*).

```{r test3-functions}
## One simulated dataset
simulate_one <- function(n_per_group, p) {
  X  <- rep(0:1, each = n_per_group)               # 0 = Lie, 1 = Palter
  M  <- rnorm(2 * n_per_group, p$b0M + p$a * X, p$sdM)
  pr <- plogis(p$b0Y + p$b * M + p$cprime * X)
  Y  <- rbinom(2 * n_per_group, 1, pr)
  data.frame(X = X, M = M, Y = Y)
}

## Analyze one dataset: a-path (OLS), b-path (logistic), indirect = a*b
analyze_one <- function(dat) {
  fa    <- lm(M ~ X, data = dat)
  a_hat <- unname(coef(fa)["X"])
  a_p   <- summary(fa)$coefficients["X", "Pr(>|t|)"]
  fb    <- glm(Y ~ X + M, family = binomial, data = dat)
  cb    <- coef(summary(fb))
  b_hat <- unname(coef(fb)["M"])
  b_p   <- cb["M", "Pr(>|z|)"]
  list(indirect  = a_hat * b_hat,
       joint_sig = (a_p < alpha) && (b_p < alpha))
}

## Power via JOINT SIGNIFICANCE (fast scan)
power_med_joint <- function(n_per_group, nSim, p) {
  hits <- replicate(nSim, {
    d <- simulate_one(n_per_group, p)
    tryCatch(analyze_one(d)$joint_sig, error = function(e) NA)
  })
  mean(hits, na.rm = TRUE)
}

## Power via NONPARAMETRIC BOOTSTRAP of a*b (faithful confirmation)
boot_indirect_ci <- function(dat, B = 999, conf = 0.95) {
  n   <- nrow(dat)
  ind <- numeric(B)
  for (j in seq_len(B)) {
    db <- dat[sample.int(n, n, replace = TRUE), ]
    ind[j] <- tryCatch({
      a  <- coef(lm(M ~ X, db))["X"]
      bb <- coef(glm(Y ~ X + M, family = binomial, db))["M"]
      a * bb
    }, error = function(e) NA_real_)
  }
  quantile(ind, c((1 - conf) / 2, 1 - (1 - conf) / 2), na.rm = TRUE)
}
power_med_boot <- function(n_per_group, nSim, p, B = 999) {
  hits <- replicate(nSim, {
    d  <- simulate_one(n_per_group, p)
    ci <- tryCatch(boot_indirect_ci(d, B = B), error = function(e) c(NA, NA))
    !any(is.na(ci)) && (ci[1] > 0 || ci[2] < 0)     # CI excludes zero
  })
  mean(hits)
}
```

Scan *N* per cell with the joint-significance proxy:

```{r test3-scan}
nSim_scan <- 1000
grid      <- c(15, 20, 25, 30, 40, 50, 75, 100)
set.seed(GLOBAL_SEED)
joint_power <- sapply(grid, function(n) power_med_joint(n, nSim_scan, pars))
test3_scan  <- data.frame(n_per_group = grid,
                          N_total     = 2 * grid,
                          joint_power = round(joint_power, 3))
test3_scan
```

```{r test3-targets}
## Smallest n/group reaching each target (NA if the grid never reaches it)
n_med <- sapply(target_powers, function(pw) {
  ok <- which(joint_power >= pw)
  if (length(ok)) grid[ok[1]] else NA_integer_
})
data.frame(power = target_powers, n_per_group = n_med)
```

Confirm the 80%-power point with the actual nonparametric bootstrap of *a·b*.
(Slow: `nSim_conf` × `B_conf` logistic fits; dial down for a quick check, or
parallelize with `parallel::mclapply` / `future.apply`.)

```{r test3-boot-confirm}
n_star    <- n_med[1]                 # 80%-power n/group from the scan
nSim_conf <- 200
B_conf    <- 999
if (!is.na(n_star)) {
  set.seed(GLOBAL_SEED)
  boot_pw <- power_med_boot(n_star, nSim = nSim_conf, p = pars, B = B_conf)
  cat(sprintf("Bootstrap-confirmed power at n = %d/group: %.3f (nSim=%d, B=%d)\n",
              n_star, boot_pw, nSim_conf, B_conf))
}
```

The mediation saturates at a small *N* and never binds the design.

# Synthesis — binding constraint and recruitment target

Required *completers* per cell = the largest requirement across the three tests.

```{r synthesis}
req <- data.frame(
  power            = target_powers,
  test1_proportion = n_baseR,           # binding
  test2_ttest      = n_t,
  test3_mediation  = n_med
)
req$binding_n_per_group <- apply(req[, c("test1_proportion", "test2_ttest",
                                         "test3_mediation")], 1,
                                 function(r) max(r, na.rm = TRUE))
req$binding_N_total <- 2 * req$binding_n_per_group
req
```


Test 1 (the proportion difference) is the binding constraint; Tests 2 and 3 are
satisfied many times over. We adopt **90% power** as the floor: the corresponding
`binding_N_total` row is the completer target, and `N_recruit` is what to launch
after inflating for ~15% exclusions.

# Sensitivity (Test 1) — a smaller-than-original effect

The proportion test is the most exposed if the true effect is smaller than the
original — for instance if this sample or platform yields a narrower palter–lie
gap. This shows how *N* grows as the true palter rate falls toward the lie rate
(held at 0.55).

```{r sensitivity}
p_palter_grid <- seq(0.71, 0.62, by = -0.01)
sens <- data.frame(
  p_palter     = p_palter_grid,
  diff_vs_lie  = round(p_palter_grid - p_lie, 2),
  cohens_h     = round(sapply(p_palter_grid, function(pp) ES.h(pp, p_lie)), 3),
  n_per_grp_80 = sapply(p_palter_grid, function(pp)
    ceiling(power.prop.test(p1 = pp, p2 = p_lie, power = 0.80, sig.level = alpha)$n)),
  n_per_grp_90 = sapply(p_palter_grid, function(pp)
    ceiling(power.prop.test(p1 = pp, p2 = p_lie, power = 0.90, sig.level = alpha)$n))
)
sens$N_total_80 <- 2 * sens$n_per_grp_80
sens$N_total_90 <- 2 * sens$n_per_grp_90
sens
```

A modest shrinkage in the palter rate inflates the requirement steeply — which is
why powering for 90% (rather than 80%) on the original effect buys a useful buffer
against a smaller true effect.

# Reproducibility

```{r session, echo=FALSE}
sessionInfo()
```
