---
title: 'Data Cleaning: Rogers et al. (2017) Study 1 Close Replication'
author: "Ke Fang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: kable
  pdf_document:
    toc: true
subtitle: "De-identify & mark duplicates, drop unused columns, then recode each variable"
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)

## --- config -----------------------------------------------------------------
in_path   <- "../data/PSYCH-251-Replication_June 8, 2026_11.17.csv"  # RAW export (contains PROLIFIC_PID)
deid_path <- "../data/data_deid.rds"                                 # de-identified, duplicate-marked
out_path  <- "../data/data_clean.rds"                                # analysis-ready (consumed by 02)

## The de-identification chunk below is the ONLY step that touches PROLIFIC_PID.
## For a public release, set this FALSE (or comment the chunk out) and ship the
## committed `data_deid.rds`; the rest of the pipeline runs unchanged.
RAW_PII_AVAILABLE <- FALSE

## --- helpers ----------------------------------------------------------------
## Parse a 7-point rating from the choice-text export. Middle points (2-6) come
## back as bare numbers; LABELED endpoints come back as text ("7 - Extremely
## deceptive"). Extract the leading integer so endpoints aren't lost to NA.
parse_rating <- function(x) as.integer(str_extract(x, "^\\s*\\d+"))

## Classify the choice by content (the honest option is identical across
## conditions; response order is counterbalanced, so do NOT key on position).
classify_choice <- function(x) {
  case_when(
    is.na(x)                                  ~ NA_character_,
    str_detect(x, "never had problems")       ~ "lie",
    str_detect(x, "drives very smoothly")     ~ "palter",
    str_detect(x, "would not start|mechanic") ~ "honest",
    TRUE                                      ~ "unclassified"
  )
}
```


# Preregistration

See: https://osf.io/w2qcd/overview

# Purpose

This is the cleaning stage. It 
(1) reads the raw Qualtrics export, marks duplicate Prolific IDs, and **discards `PROLIFIC_PID`**, saving a de-identified intermediate; then 
(2) drops irrelevant columns and recodes each analysis variable in its own subsection. 
The output (`data_clean.rds`) carries the variables `02_analysis.Rmd` expects, the
eligibility/exclusion flags, the two free-text responses, and several
quality/fraud-check variables (reCAPTCHA, Qualtrics' duplicate flag, the
AI free response) for inspecting suspicious responses.

Every response is retained — including incompletes/non-finishers, 
comprehension screen-outs, and duplicates — but *marked* with flags 
(`is_real`, `passed_comprehension`, `has_choice`, `ai_excluded`, `is_duplicate`,
`analysis_ok`, and the quality variables). No rows are dropped. All
sample selection (the preregistered sample and any exploratory subsets)
happens at the start of `02_analysis.Rmd` by filtering on these flags.

`PROLIFIC_PID` never leaves the de-identification step and is not stored
in any saved file.

# De-identification and duplicate marking

> **PII step — run once on the raw export, then set
> `RAW_PII_AVAILABLE <- FALSE` (or comment this chunk out) for any
> public release.** It is the only code that reads `PROLIFIC_PID`. It
> marks duplicates (keeping the first *complete, eligible* response per
> Prolific ID), drops the ID, and writes the de-identified
> `data_deid.rds` that every later step reads.

```{r deidentify, eval=RAW_PII_AVAILABLE}
raw0 <- read_csv(in_path, col_names = FALSE,
                 col_types = cols(.default = col_character()))
import_ids <- str_match(unlist(raw0[3, ]), '"ImportId":"([^"]+)"')[, 2]
import_ids[is.na(import_ids)] <- as.character(unlist(raw0[1, ]))[is.na(import_ids)]

d_raw <- raw0 |>
  slice(-(1:3)) |>
  set_names(import_ids) |>
  transmute(
    ## identifiers & metadata
    response_id = `_recordId`,            # Qualtrics response id (not PII)
    prolific_id = PROLIFIC_PID,           # PII -- used here, then dropped
    start_date  = startDate,
    status, finished, progress, duration,

    ## quality / fraud indicators (retained for inspection; not prereg criteria)
    recaptcha          = Q_RecaptchaScore,
    qualtrics_dup_flag = Q_DuplicateRespondent,

    ## assignment & randomization
    consent          = QID3,
    condition_raw    = condition,
    text_order_raw   = text_order,
    rating_bonus_raw = rating_bonus_order,
    bonus_draw_raw   = bonus_draw,

    ## choice (palter block / lie block)
    choice_palter = QID5,  choice_lie = QID7,

    ## ratings
    decep_palter   = QID13, decep_lie   = QID17,
    honest_palter  = QID15, honest_lie  = QID18,
    ethical_palter = QID16, ethical_lie = QID19,

    ## comprehension
    comp_goal_palter  = QID26, comp_goal_lie  = QID29,
    comp_false_palter = QID27, comp_false_lie = QID36,

    ## probability memory (honest-response prob / misleading-response prob)
    memhon_palter = QID28, memhon_lie = QID38,
    memmis_palter = QID39, memmis_lie = QID31,

    ## demographics & free text
    age_raw    = QID24_TEXT,
    gender_raw = QID21,
    ai_text    = QID70_TEXT,   # AI-use disclosure (free text)
    strategy   = QID69_TEXT    # open-ended reasoning (free text)
  ) |>
  mutate(across(everything(), ~ na_if(.x, "")))

## Mark duplicates: keep the FIRST complete eligible response per Prolific ID
## (prefer made-a-choice, then finished, then earliest start), drop the ID.
d_deid <- d_raw |>
  arrange(prolific_id,
          desc(!is.na(choice_palter) | !is.na(choice_lie)),
          desc(finished %in% c("True", "TRUE", "1")),
          start_date) |>
  group_by(prolific_id) |>
  mutate(dup_rank = if_else(is.na(prolific_id), 1L, row_number())) |>
  ungroup() |>
  mutate(is_duplicate = !is.na(prolific_id) & dup_rank > 1) |>
  select(-prolific_id, -dup_rank)            # <-- PROLIFIC_PID dropped here

saveRDS(d_deid, deid_path)
cat("Wrote", deid_path, "-", nrow(d_deid), "rows;",
    sum(d_deid$is_duplicate), "duplicate(s) marked; PROLIFIC_PID dropped.\n")
```

# Load de-identified data

Everything below reads the de-identified file, so it works whether or
not the PII step above ran.

```{r load-deid}
d <- readRDS(deid_path)
cat("Loaded", nrow(d), "rows x", ncol(d), "cols (no PROLIFIC_PID).\n")
cat("'PROLIFIC_PID' present in loaded data:", "PROLIFIC_PID" %in% names(d),
    "| 'prolific_id' present:", "prolific_id" %in% names(d), "\n")
glimpse(d)
```

# Variable cleaning

## Response validity

Real = non-preview and finished. (Comprehension screen-outs are
sometimes marked `finished = True`, so the analysis sample is defined by
the comprehension and choice flags below, not by `is_real` alone.)

```{r v-valid}
d <- d |>
  mutate(is_real  = !str_detect(coalesce(status, ""), "Preview") &
                    finished %in% c("True", "TRUE", "1"),
         progress = as.integer(progress),
         duration = as.integer(duration))
d |> count(status, finished, is_real)
```

## Consent

```{r v-consent}
d <- d |> mutate(consent_ok = coalesce(consent == "Yes", FALSE))
d |> count(consent, consent_ok)
```

## Condition

```{r v-condition}
d <- d |>
  mutate(condition     = factor(condition_raw,
                                levels = c("lying_truth", "paltering_truth"),
                                labels = c("lie", "palter")),
         condition_num = if_else(condition == "palter", 1L, 0L))
d |> count(condition_raw, condition, condition_num)
```

## Choice → `misleading`

```{r v-choice}
d <- d |>
  mutate(choice_text = coalesce(choice_palter, choice_lie),
         choice      = classify_choice(choice_text),
         misleading  = case_when(choice %in% c("palter", "lie") ~ 1L,
                                 choice == "honest"             ~ 0L,
                                 TRUE                            ~ NA_integer_))
d |> count(condition, choice)
```

## Ratings: deceptiveness, honesty, ethicality

```{r v-ratings}
## the raw endpoint-label issue, before fixing:
d |> transmute(v = coalesce(decep_palter, decep_lie)) |> drop_na() |>
  count(v) |> arrange(v) |> print()

d <- d |>
  mutate(deceptive = parse_rating(coalesce(decep_palter, decep_lie)),
         honest    = parse_rating(coalesce(honest_palter, honest_lie)),
         ethical   = parse_rating(coalesce(ethical_palter, ethical_lie)))

d |> summarise(across(c(deceptive, honest, ethical),
                      list(min = ~min(.x, na.rm = TRUE), max = ~max(.x, na.rm = TRUE),
                           n_na = ~sum(is.na(.x)))))
```

## Reverse-coded deceptiveness and ethicality index

```{r v-index}
d <- d |>
  mutate(decep_rev    = 8L - deceptive,
         ethics_index = rowMeans(cbind(honest, ethical, decep_rev)))  # NA if any missing
d |> filter(!is.na(ethics_index)) |> group_by(condition) |>
  summarise(n = n(), mean_index = round(mean(ethics_index), 2), .groups = "drop")
```

## Comprehension screen

```{r v-comprehension}
d <- d |>
  mutate(comp1_ok = str_detect(coalesce(comp_goal_palter,  comp_goal_lie),  "To sell the car"),
         comp2_ok = str_detect(coalesce(comp_false_palter, comp_false_lie), "two small dents"),
         passed_comprehension = coalesce(comp1_ok, FALSE) & coalesce(comp2_ok, FALSE))
d |> count(comp1_ok, comp2_ok, passed_comprehension)
```

## Probability memory (exploratory)

Correct answer for the *misleading*-response item is condition-dependent
(60% palter / 80% lie); the honest-response item is 30% in both. Not an
exclusion.

```{r v-memory}
d <- d |>
  mutate(mem_misleading = coalesce(memmis_palter, memmis_lie),
         mem_honest     = coalesce(memhon_palter, memhon_lie),
         mem_misleading_correct = case_when(
           is.na(mem_misleading)                           ~ NA,
           condition == "palter" & mem_misleading == "60%" ~ TRUE,
           condition == "lie"    & mem_misleading == "80%" ~ TRUE,
           TRUE                                            ~ FALSE),
         mem_honest_correct = case_when(
           is.na(mem_honest)   ~ NA,
           mem_honest == "30%" ~ TRUE,
           TRUE                ~ FALSE))
d |> group_by(condition) |>
  summarise(recall_mis = mean(mem_misleading_correct, na.rm = TRUE),
            recall_hon = mean(mem_honest_correct, na.rm = TRUE), .groups = "drop")
```

## Bonus draw and bonus earned

Earned the \$1 if the 1–100 draw ≤ the success probability of the chosen
response (30 honest / 60 palter / 80 lie). Descriptive; earned rates
should track those.

```{r v-bonus}
d <- d |>
  mutate(bonus_draw      = as.integer(bonus_draw_raw),
         bonus_threshold = case_when(choice == "honest" ~ 30L,
                                     choice == "palter" ~ 60L,
                                     choice == "lie"    ~ 80L,
                                     TRUE                ~ NA_integer_),
         bonus_earned    = bonus_draw <= bonus_threshold)
d |> filter(!is.na(choice)) |> group_by(choice) |>
  summarise(n = n(), earned_rate = round(mean(bonus_earned, na.rm = TRUE), 2),
            draw_mean = round(mean(bonus_draw, na.rm = TRUE), 1), .groups = "drop")
```

## Order variables

```{r v-order}
d <- d |>
  mutate(text_order         = factor(text_order_raw),
         rating_bonus_order = factor(rating_bonus_raw,
                                     levels = c("ratings_first", "bonus_first")))
d |> count(rating_bonus_order)
```

## Demographics

```{r v-demographics}
d <- d |>
  mutate(age    = suppressWarnings(as.integer(na_if(str_trim(age_raw), "-99"))),
         gender = gender_raw)
d |> summarise(age_min = min(age, na.rm = TRUE), age_med = median(age, na.rm = TRUE),
               age_max = max(age, na.rm = TRUE), n_age_na = sum(is.na(age)))
d |> count(gender)
```

## Free-text responses

Keep both open-ended responses for qualitative review (and the AI screen
below). Qualtrics writes `-99` for some skipped text fields; treat those
as missing.

```{r v-freetext}
d <- d |>
  mutate(strategy = na_if(str_trim(strategy), "-99"),
         ai_text  = na_if(str_trim(ai_text),  "-99"))
d |> summarise(n_strategy = sum(!is.na(strategy)), n_ai_text = sum(!is.na(ai_text)))
```

## AI-use disclosure

Per the preregistration, exclude only participants who **explicitly
disclose** AI use. A keyword screen flags candidates, but most are
emphatic denials, so we exclude only those whose text affirmatively
indicates AI use. All flagged responses are printed for manual
verification; override with `ai_force_exclude` / `ai_force_keep` (by
`response_id`) if any are misclassified.

```{r v-ai}
ai_terms  <- "chatgpt|chat gpt|\\bgpt\\b|openai|\\bclaude\\b|gemini|copilot|\\bbard\\b|\\bllm\\b|deepseek|artificial intelligence|\\bai\\b"
ai_affirm <- "^\\s*(yes|si|yeah|yep|sure)\\b|\\b(used an? ai|used chatgpt|used gpt|using ai|with chatgpt|with ai)\\b"

ai_force_exclude <- character(0)   # response_ids to FORCE exclude after review
ai_force_keep    <- character(0)   # response_ids to FORCE keep after review

d <- d |>
  mutate(ai_text_l   = str_to_lower(coalesce(ai_text, "")),
         ai_flagged  = str_detect(ai_text_l, ai_terms),
         ai_excluded = (ai_flagged & str_detect(ai_text_l, ai_affirm) &
                          !response_id %in% ai_force_keep) |
                         response_id %in% ai_force_exclude) |>
  select(-ai_text_l)

cat("AI: flagged for review =", sum(d$ai_flagged),
    "| excluded as disclosures =", sum(d$ai_excluded), "\n\n")
d |> filter(ai_flagged) |> transmute(ai_excluded, response_id, ai_text) |>
  arrange(desc(ai_excluded)) |> print(n = Inf)
```

## Quality / fraud indicators (retained, not exclusion criteria)

reCAPTCHA v3 score (higher ≈ more human) and Qualtrics' own duplicate
flag are **not** preregistered exclusion criteria, so they do not feed
`analysis_ok`. We retain them so the team can inspect suspicious
("wired") responses; if you later decide to act on them, they are in the
saved data.

```{r v-quality}
d <- d |>
  mutate(recaptcha          = suppressWarnings(as.numeric(recaptcha)),
         qualtrics_dup_flag  = coalesce(str_to_lower(qualtrics_dup_flag) == "true", FALSE))
d |> summarise(recaptcha_min = min(recaptcha, na.rm = TRUE),
               recaptcha_lt_0.5 = sum(recaptcha < 0.5, na.rm = TRUE),
               qualtrics_dup_flagged = sum(qualtrics_dup_flag),
               our_dup_flagged = sum(is_duplicate))
```

## Duplicates (marked in the de-identification step)

`is_duplicate` was set above from `PROLIFIC_PID` (now discarded). Note
this is distinct from Qualtrics' fingerprint-based `qualtrics_dup_flag`.

```{r v-duplicates}
d |> count(is_duplicate)
```

# Eligibility flag and attrition summary

`analysis_ok` uses only the preregistered criteria (consent, real,
randomized, comprehension, valid choice, no explicit AI disclosure, not
a Prolific-ID duplicate). reCAPTCHA and the Qualtrics duplicate flag are
deliberately excluded from this rule. The flag is *added to every row*;
no rows are dropped here. The table below is a descriptive attrition
summary — the saved file retains all responses, and `02` applies the
flag.

```{r analysis-ok}
d <- d |>
  mutate(has_choice  = !is.na(misleading),
         analysis_ok = consent_ok & is_real & !is.na(condition) &
                       passed_comprehension & has_choice &
                       !ai_excluded & !is_duplicate)

tibble(
  step = c("recruited", "real (finished, non-preview)", "consented",
           "randomized to a condition", "passed comprehension",
           "made a valid choice", "not an AI disclosure", "not a duplicate"),
  remaining = c(
    nrow(d),
    sum(d$is_real),
    sum(d$is_real & d$consent_ok),
    sum(d$is_real & d$consent_ok & !is.na(d$condition)),
    sum(d$is_real & d$consent_ok & !is.na(d$condition) & d$passed_comprehension),
    sum(d$is_real & d$consent_ok & !is.na(d$condition) & d$passed_comprehension & d$has_choice),
    sum(d$is_real & d$consent_ok & !is.na(d$condition) & d$passed_comprehension & d$has_choice & !d$ai_excluded),
    sum(d$analysis_ok))) |>
  mutate(excluded_here = lag(remaining) - remaining)
```

# Integrity checks

```{r integrity}
elig <- d |> filter(analysis_ok)
cat("ANALYSIS-ELIGIBLE:", nrow(elig), "\n\n")
cat("condition balance:\n");           print(count(elig, condition))
cat("\nrating/bonus order balance:\n"); print(count(elig, rating_bonus_order))
cat("\nunclassified choices (eligible):", sum(elig$choice == "unclassified", na.rm = TRUE), "\n")
cat("choice/condition contamination:",
    sum((elig$condition == "palter" & elig$choice == "lie") |
        (elig$condition == "lie"    & elig$choice == "palter"), na.rm = TRUE), "\n")
```

# Save the full flagged dataset

Select the analysis contract, the eligibility/exclusion flags, both
free-text responses, and the quality-check variables — keeping **all
rows** (incompletes, screen-outs, and duplicates included; they are
flagged, not removed). **No `PROLIFIC_PID`/`prolific_id`.** `02` creates
the analysis samples from the flags.

```{r save}
data_clean <- d |>
  select(response_id,
         condition, condition_num,
         choice, misleading,
         deceptive, honest, ethical, decep_rev, ethics_index,
         comp1_ok, comp2_ok, passed_comprehension,
         mem_misleading, mem_honest, mem_misleading_correct, mem_honest_correct,
         bonus_draw, bonus_earned,
         text_order, rating_bonus_order,
         age, gender,
         strategy, ai_text,                                   # free-text responses
         recaptcha, qualtrics_dup_flag,                       # quality / fraud checks
         consent_ok, is_real, ai_flagged, ai_excluded, is_duplicate, duration,
         analysis_ok)

## guards
stopifnot(!any(c("prolific_id", "PROLIFIC_PID") %in% names(data_clean)))
required <- c("misleading","condition","condition_num","ethics_index",
              "honest","ethical","deceptive","decep_rev","rating_bonus_order",
              "mem_misleading_correct","mem_honest_correct","bonus_earned",
              "analysis_ok","age","gender")
stopifnot(all(required %in% names(data_clean)))

saveRDS(data_clean, out_path)
cat("Saved ALL", nrow(data_clean), "rows to", out_path, "(no PROLIFIC_PID);",
    sum(data_clean$analysis_ok), "flagged analysis-eligible,",
    nrow(data_clean) - sum(data_clean$analysis_ok), "flagged out (retained, not dropped).\n")
```

# Reproducibility

```{r session, echo=FALSE}
sessionInfo()
```
