---
title: "Analysis: Rogers et al. (2017) Study 1 Close Replication"
author: "Ke Fang"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: flatly
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = FALSE)

library(tidyverse)
library(lavaan)      # mediation (WLSMV / probit)
library(semTools)    # >= 0.5-6 for monteCarloCI()

set.seed(42)

## --- config -----------------------------------------------------------------
data_path     <- "../data/data_clean.rds"   # produced by 01_import_preprocess.R
N_MC          <- 20000               # Monte Carlo draws for indirect-effect CI
B_BOOT        <- 2000                # bootstrap resamples for the indirect effect
```

# Purpose and structure

This document follows a **models-first** design. Each confirmatory test and the
mediation are written as a *function of a data frame*; the analysis samples are
defined separately as a named list; and a thin driver runs every model on every
sample. Adding a new (e.g. exploratory) sample is a single line in the `samples`
list, and all samples are guaranteed to pass through identical code.

It assumes you have already run `01_import_preprocess.R`, which writes the
analysis-ready data (`data_clean.rds`) with the variables specified in the
contract below. Confirmatory analyses subset on `analysis_ok`.

# Variable contract

Working backwards from each analysis, these are the variables every model needs,
their required form, and where they come from. The preprocessing script produces
all of them.

| Variable | Type / form | Derived from | Consumed by |
|---|---|---|---|
| `misleading` | integer 0/1 (declarable `ordered`) | `coalesce(QID5, QID7)` classified by content | H1 outcome; mediation Y; memory analysis covariate |
| `condition` | factor (`lie`,`palter`) | embedded `condition` | H1 & H2 grouping; descriptives |
| `condition_num` | numeric 0/1 (lie=0, palter=1) | `condition` | mediation X (lavaan needs numeric) |
| `honest`, `ethical`, `deceptive` | integer 1–7 | `parse_rating(coalesce(...))` | reliability (α); item-level contingency; index |
| `decep_rev` | integer 1–7 | `8 - deceptive` | index; reliability; contingency |
| `ethics_index` | numeric (mean of 3) | `mean(honest, ethical, decep_rev)` | H2 outcome; mediation M |
| `rating_bonus_order` | factor | embedded | order exploratory; ratings-first robustness sample |
| `mem_misleading_correct`, `mem_honest_correct` | logical | memory items vs condition-specific key | recall-pass sample; memory exploratory |
| `bonus_earned` | logical | `bonus_draw <= threshold(choice)` | incentive integrity (descriptive) |
| `analysis_ok` | logical | consent/real/comprehension/choice/AI/dup | defines confirmatory sample |
| `age`, `gender` | int / chr | demographics | sample description |

```{r load-and-check}
clean <- readRDS(data_path)

required <- c("misleading","condition","condition_num","ethics_index",
              "honest","ethical","deceptive","decep_rev",
              "rating_bonus_order","mem_misleading_correct","mem_honest_correct",
              "bonus_earned","analysis_ok","age","gender")
missing_vars <- setdiff(required, names(clean))
if (length(missing_vars))
  stop("Missing required variables (re-run 01_import_preprocess.R): ",
       paste(missing_vars, collapse = ", "))

clean <- clean |> mutate(condition = factor(condition, levels = c("lie","palter")))
cat("Loaded the FULL flagged dataset:", nrow(clean), "rows (",
    sum(clean$analysis_ok), "flagged eligible,",
    nrow(clean) - sum(clean$analysis_ok), "flagged out but retained ).\n")
```

# Analysis samples (defined up front)

`data_clean.rds` is the full flagged dataset — `01` drops no rows, so it contains
all responses (incompletes, screen-outs, and duplicates included, each carrying
flags). We build every analysis sample here, at the start, by filtering on those
flags. The confirmatory base is `analysis_ok`; add an exploratory subset by
appending one line and every model below picks it up automatically.

```{r samples}
base <- clean |> filter(analysis_ok)

samples <- list(
  preregistered = base,                                                  # confirmatory
  recall_pass   = base |> filter(mem_misleading_correct, mem_honest_correct),
  ratings_first = base |> filter(rating_bonus_order == "ratings_first"),  # original-analog
  quality_clean = base |> filter(recaptcha >= 0.5, !qualtrics_dup_flag)
)

tibble(sample = names(samples), n = map_int(samples, nrow))
```

# Models (defined once)

Small effect-size helpers first, then one function per analysis. Each takes a
data frame (already subset to a sample) and returns a tidy one-row (or few-row)
tibble, so results stack cleanly across samples.

```{r helpers}
cohens_h <- function(p1, p2) 2*asin(sqrt(p1)) - 2*asin(sqrt(p2))

cohens_d <- function(x1, x2) {            # x1 - x2, pooled SD
  n1 <- sum(!is.na(x1)); n2 <- sum(!is.na(x2))
  sp <- sqrt(((n1-1)*var(x1, na.rm=TRUE) + (n2-1)*var(x2, na.rm=TRUE)) / (n1+n2-2))
  (mean(x1, na.rm=TRUE) - mean(x2, na.rm=TRUE)) / sp
}

cronbach_alpha <- function(items) {       # items: data frame of columns
  items <- na.omit(as.matrix(items)); k <- ncol(items); vc <- var(items)
  (k/(k-1)) * (1 - sum(diag(vc))/sum(vc))
}

odds_ratio <- function(tab) {             # tab: rows c(lie,palter) x cols c(0,1)
  if (any(tab == 0)) tab <- tab + 0.5     # Haldane correction if a zero cell
  or <- (tab["palter","1"]*tab["lie","0"]) / (tab["palter","0"]*tab["lie","1"])
  se <- sqrt(sum(1/tab))
  c(OR = unname(or),
    OR_lo = unname(exp(log(or) - 1.96*se)),
    OR_hi = unname(exp(log(or) + 1.96*se)))
}

## robust pull of a Monte Carlo CI for one defined parameter
mc_ci <- function(fit, label, nRep, seed = 42) {
  set.seed(seed)
  mc <- tryCatch(semTools::monteCarloCI(fit, nRep = nRep), error = function(e) NULL)
  if (is.null(mc) || !(label %in% rownames(mc))) return(c(lo = NA_real_, hi = NA_real_))
  cols <- colnames(mc)
  lo <- cols[grepl("lower", cols, ignore.case = TRUE)][1]
  hi <- cols[grepl("upper", cols, ignore.case = TRUE)][1]
  c(lo = mc[label, lo], hi = mc[label, hi])
}

## robust pull of a parameter row by label OR defined-name
get_par <- function(pe, name) {
  row <- pe[!is.na(pe$label) & pe$label == name, , drop = FALSE]
  if (nrow(row) == 0) row <- pe[pe$op == ":=" & pe$lhs == name, , drop = FALSE]
  row
}
```

## H1 — choice rate

```{r model-h1}
run_h1 <- function(d) {
  d <- d |> filter(!is.na(misleading), !is.na(condition))
  tab <- table(factor(d$condition, levels = c("lie","palter")),
               factor(d$misleading, levels = c(0, 1)))
  xp <- d$misleading[d$condition == "palter"]
  xl <- d$misleading[d$condition == "lie"]
  chi <- suppressWarnings(chisq.test(tab, correct = FALSE))   # Pearson, no Yates
  pt  <- suppressWarnings(prop.test(c(sum(xp), sum(xl)), c(length(xp), length(xl)),
                                    correct = FALSE))          # CI for palter - lie
  orr <- odds_ratio(tab)
  tibble(n = nrow(d), n_palter = length(xp), n_lie = length(xl),
         p_palter = mean(xp), p_lie = mean(xl), diff = mean(xp) - mean(xl),
         diff_lo = pt$conf.int[1], diff_hi = pt$conf.int[2],
         cohens_h = cohens_h(mean(xp), mean(xl)),
         OR = orr["OR"], OR_lo = orr["OR_lo"], OR_hi = orr["OR_hi"],
         chisq = unname(chi$statistic), df = unname(chi$parameter), p = chi$p.value)
}
```

## H2 — ethicality index (+ reliability gate)

```{r model-h2}
run_h2 <- function(d) {
  d <- d |> filter(!is.na(ethics_index), !is.na(condition))
  xp <- d$ethics_index[d$condition == "palter"]
  xl <- d$ethics_index[d$condition == "lie"]
  tt <- t.test(xp, xl)                      # Welch; estimate/CI for palter - lie
  alpha_overall <- cronbach_alpha(d |> select(honest, ethical, decep_rev))
  tibble(n = nrow(d), n_palter = length(xp), n_lie = length(xl),
         m_palter = mean(xp), sd_palter = sd(xp),
         m_lie = mean(xl), sd_lie = sd(xl),
         diff = mean(xp) - mean(xl), diff_lo = tt$conf.int[1], diff_hi = tt$conf.int[2],
         cohens_d = cohens_d(xp, xl),
         t = unname(tt$statistic), df = unname(tt$parameter), p = tt$p.value,
         alpha_overall = alpha_overall)
}

run_reliability <- function(d) {
  items <- d |> select(honest, ethical, decep_rev)
  by <- d |> group_by(condition) |>
    summarise(alpha = cronbach_alpha(pick(honest, ethical, decep_rev)),
              n = sum(complete.cases(pick(honest, ethical, decep_rev))), .groups = "drop")
  list(alpha_overall = cronbach_alpha(items),
       by_condition = by,
       inter_item = round(cor(na.omit(items)), 2))
}
```

## Mediation — condition → ethicality → choice

Primary: the single-index mediator with the binary outcome declared `ordered`
(WLSMV / probit). Indirect effect via Monte Carlo CI (primary) and, optionally,
a bias-corrected bootstrap (sensitivity). Probit coefficients are not comparable
in magnitude to the original's logit values; the sign and significance of the
indirect effect are the replication target, and we report the standardized
indirect effect.

```{r model-mediation}
MED_MODEL <- '
  ethics_index ~ a*condition_num
  misleading   ~ b*ethics_index + cp*condition_num
  indirect := a*b
  total    := indirect + cp
'

run_mediation_index <- function(d, n_mc = N_MC, B = B_BOOT, seed = 42) {
  d <- d |> filter(!is.na(ethics_index), !is.na(misleading), !is.na(condition_num))
  fit <- tryCatch(sem(MED_MODEL, data = d, ordered = "misleading",
                      estimator = "WLSMV"), error = function(e) NULL)
  if (is.null(fit))
    return(tibble(n = nrow(d), note = "point-estimate model did not fit"))

  pe  <- parameterEstimates(fit, standardized = TRUE)
  a   <- get_par(pe, "a"); b <- get_par(pe, "b")
  cp  <- get_par(pe, "cp"); ind <- get_par(pe, "indirect")
  mc  <- mc_ci(fit, "indirect", n_mc, seed)

  ## Bias-corrected bootstrap of the indirect effect -- ALWAYS run (the analog of
  ## the original's Sobel-Goodman bootstrap). NB: lavaan refuses se="bootstrap"
  ## with WLSMV ("use (D)WLS estimator for bootstrap"), so the bootstrap fit uses
  ## DWLS -- which gives IDENTICAL point estimates (the "MV" only adjusts the
  ## analytic test statistic / SEs, which the bootstrap replaces with the resample
  ## distribution). On some resamples bca.simple can be unavailable for a defined
  ## parameter, so we fall back to percentile and never fail silently. boot_n =
  ## successful resamples, boot_type = which CI was returned, boot_note = the
  ## error message if the bootstrap fit itself failed.
  set.seed(seed)
  fitb <- tryCatch(sem(MED_MODEL, data = d, ordered = "misleading",
                       estimator = "DWLS", se = "bootstrap", bootstrap = B),
                   error = function(e) conditionMessage(e))
  boot_lo <- NA_real_; boot_hi <- NA_real_
  boot_n  <- NA_integer_; boot_type <- NA_character_; boot_note <- NA_character_
  if (inherits(fitb, "lavaan")) {
    boot_n <- tryCatch(nrow(lavInspect(fitb, "boot")), error = function(e) NA_integer_)
    if (is.na(boot_n))
      boot_n <- tryCatch(nrow(fitb@boot$coef), error = function(e) NA_integer_)
    bi <- tryCatch(get_par(parameterEstimates(fitb, boot.ci.type = "bca.simple"), "indirect"),
                   error = function(e) NULL)
    if (!is.null(bi) && is.finite(bi$ci.lower) && is.finite(bi$ci.upper)) {
      boot_lo <- bi$ci.lower; boot_hi <- bi$ci.upper; boot_type <- "bca.simple"
    } else {                                   # robust fallback
      bp <- get_par(parameterEstimates(fitb, boot.ci.type = "perc"), "indirect")
      boot_lo <- bp$ci.lower; boot_hi <- bp$ci.upper; boot_type <- "perc"
    }
  } else {
    boot_note <- substr(as.character(fitb), 1, 80)   # the captured error message
  }

  tibble(n = nrow(d),
         a = a$est, b = b$est, cprime = cp$est,
         indirect = ind$est, indirect_std = ind$std.all,
         mc_lo = mc["lo"], mc_hi = mc["hi"],
         mc_excl_0 = !is.na(mc["lo"]) && (mc["lo"] > 0 | mc["hi"] < 0),
         boot_lo = boot_lo, boot_hi = boot_hi,
         boot_excl_0 = !is.na(boot_lo) && (boot_lo > 0 | boot_hi < 0),
         boot_n = boot_n, boot_type = boot_type, boot_note = boot_note)
}

## Contingency (use only if overall alpha < .70): three SEPARATE single-mediator
## models, Bonferroni-adjusted, interpreted exploratorily. No simultaneous
## (parallel) model -> no collinearity among mediators.
run_mediation_items <- function(d, n_mc = N_MC, seed = 42) {
  items <- c("honest", "ethical", "decep_rev")
  d <- d |> filter(!is.na(misleading), !is.na(condition_num),
                   !is.na(honest), !is.na(ethical), !is.na(decep_rev))
  map_dfr(items, function(it) {
    f <- sprintf('%1$s ~ a*condition_num\n misleading ~ b*%1$s + cp*condition_num\n indirect := a*b', it)
    fit <- tryCatch(sem(f, data = d, ordered = "misleading", estimator = "WLSMV"),
                    error = function(e) NULL)
    if (is.null(fit)) return(tibble(mediator = it, note = "model did not fit"))
    ind <- get_par(parameterEstimates(fit, standardized = TRUE), "indirect")
    mc  <- mc_ci(fit, "indirect", n_mc, seed)
    tibble(mediator = it, indirect = ind$est, indirect_std = ind$std.all,
           mc_lo = mc["lo"], mc_hi = mc["hi"])
  }) |> mutate(bonferroni_alpha = 0.05 / length(items))
}
```

## Descriptives / manipulation checks

```{r model-describe}
describe_sample <- function(d) {
  d |> summarise(
    n = n(), n_palter = sum(condition == "palter"), n_lie = sum(condition == "lie"),
    mislead_palter = mean(misleading[condition == "palter"]),
    mislead_lie    = mean(misleading[condition == "lie"]),
    bonus_earned_rate = mean(bonus_earned, na.rm = TRUE),
    recall_mis_correct = mean(mem_misleading_correct, na.rm = TRUE),
    recall_hon_correct = mean(mem_honest_correct, na.rm = TRUE),
    mean_age = mean(age, na.rm = TRUE)
  )
}
```

# Sample descriptions

```{r run-describe}
imap_dfr(samples, ~ describe_sample(.x) |> mutate(sample = .y, .before = 1))
```

# H1 — choice rate, across samples

```{r run-h1}
imap_dfr(samples, ~ run_h1(.x) |> mutate(sample = .y, .before = 1)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

The confirmatory test is the `preregistered` row: support requires *p* < .05 and
`p_palter > p_lie`. Other rows show robustness of the same effect under
alternative sample definitions.

# H2 — ethicality index, across samples

```{r run-h2}
imap_dfr(samples, ~ run_h2(.x) |> mutate(sample = .y, .before = 1)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

`alpha_overall` is the reliability gate. If it is ≥ .70 in the `preregistered`
sample, H2 and the mediation use the index (the rows above and below). If it is
< .70, switch to the item-level contingency (below). The `ratings_first` row is
the pre-specified robustness check matching the original's fixed order.

```{r run-reliability}
rel <- run_reliability(samples$preregistered)
cat("Preregistered-sample overall alpha:", round(rel$alpha_overall, 3),
    if (rel$alpha_overall >= .70) " (>= .70: use index)\n" else " (< .70: use contingency)\n")
cat("\nBy condition:\n"); print(rel$by_condition)
cat("\nInter-item correlations:\n"); print(rel$inter_item)
```

# Mediation, across samples

The indirect effect of condition on the choice to mislead, through perceived
ethicality. The single-index model is fit per sample with the binary outcome
declared `ordered` (WLSMV / probit). For each sample we report both the Monte
Carlo CI (primary) and the bias-corrected bootstrap CI (the analog of the
original's Sobel-Goodman bootstrap); the bootstrap now runs on every sample. This
is the slow step — the model is refit `B_BOOT` times per sample (under DWLS, which
lavaan requires for bootstrapping and which gives the same point estimates as
WLSMV) — so expect it to take a while.

```{r run-mediation}
imap_dfr(samples, ~ run_mediation_index(.x) |> mutate(sample = .y, .before = 1)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

`mc_excl_0` / `boot_excl_0 == TRUE` mean the 95% Monte Carlo / bootstrap CI for
the indirect effect excludes zero (the inference criterion); the two should agree.
`indirect_std` is the standardized indirect effect (the magnitude to report,
given the probit scale). `boot_n` is the number of successful bootstrap resamples
(below `B_BOOT` means some resamples failed to converge), `boot_type` is which CI
was returned (`bca.simple`, or `perc` if BCa was unavailable for that fit), and
`boot_note` carries the error message if the bootstrap fit itself failed.

## Contingency: item-level mediation (only if overall alpha < .70)

```{r run-mediation-items}
if (rel$alpha_overall < .70) {
  cat("Overall alpha < .70 -> running three single-mediator models (Bonferroni):\n")
  print(imap_dfr(samples, ~ run_mediation_items(.x) |> mutate(sample = .y, .before = 1)) |>
          mutate(across(where(is.numeric), ~ round(.x, 3))))
} else {
  cat("Overall alpha >= .70 in the preregistered sample; index mediation is used.\n",
      "The item-level contingency is not triggered. (run_mediation_items() is\n",
      "available if you want it as a labeled exploratory analysis.)\n")
}
```

# Exploratory: rating/bonus order effect on ratings

Within the preregistered sample, does seeing the bonus outcome before vs. after
the ratings shift the ethicality index? (Descriptive + a 2×2 condition × order
model on the index.)

```{r run-order}
ord <- samples$preregistered
ord |> group_by(condition, rating_bonus_order) |>
  summarise(n = n(), mean_index = round(mean(ethics_index, na.rm = TRUE), 2),
            .groups = "drop")

summary(lm(ethics_index ~ condition * rating_bonus_order, data = ord))
```

# Reproducibility

```{r session, echo=FALSE}
sessionInfo()
```
