---
title: 'Psych 251 PS4: Simulation + Analysis'
author: "Mike Frank"
date: "2019"
format: 
  html:
    toc: true
---

This is problem set #4, in which we want you to integrate your knowledge of data wrangling with some basic simulation skills. It's a short problem set to help consolidate your `ggplot2` skills and then help you get your feet wet in testing statistical concepts through "making up data" rather than consulting a textbook or doing math. 

For ease of reading, please separate your answers from our text by marking our text with the `>` character (indicating quotes). 

# Part 1: ggplot practice

This part is a warmup, it should be relatively straightforward `ggplot2` practice.

Load data from Frank, Vul, Saxe (2011, Infancy), a study in which we measured infants' looking to hands in moving scenes. There were infants from 3 months all the way to about two years, and there were two movie conditions (`Faces_Medium`, in which kids played on a white background, and `Faces_Plus`, in which the backgrounds were more complex and the people in the videos were both kids and adults). An eye-tracker measured children's attention to faces. This version of the dataset only gives two conditions and only shows the amount of looking at hands (other variables were measured as well). 

```{r}
library(tidyverse)
fvs <- read_csv("data/FVS2011-hands.csv")
```

First, use `ggplot` to plot a histogram of the ages of children in the study. NOTE: this is a repeated measures design, so you can't just take a histogram of every measurement. 

```{r}
# Assuming there's an identifier for each child, select distinct ages
unique_ages <- fvs %>% 
  distinct(subid, .keep_all = TRUE) %>% 
  select(age)

# Plot the histogram using ggplot
ggplot(unique_ages, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Histogram of Ages", x = "Age (in months)", y = "Count of Children") +
  theme_minimal()
```

Second, make a scatter plot showing hand looking as a function of age and condition. Add appropriate smoothing lines. Take the time to fix the axis labels and make the plot look nice.

```{r}
ggplot(fvs, aes(x = age, y = hand.look, color = condition)) +
  geom_point(alpha = 0.6) + # Set transparency to see overlapping points
  geom_smooth(method = "lm", se = TRUE) + # Add linear smoothing lines without confidence intervals
  labs(title = "Hand Looking as a Function of Age and Condition",
       x = "Age (in months)",
       y = "Proportion of Time Looking at Hands",
       color = "Condition") +
  scale_color_brewer(palette = "Set1") + # Use a color palette that's distinct
  theme_minimal() + # Minimal theme for a nicer look
  theme(legend.position = "bottom") # Move legend to the bottom

```

What do you conclude from this pattern of data?

> Overall, children spend longer times looking at hands as they get older, and that this is especially pronounced when looking at hands in context. 

What statistical analyses would you perform here to quantify these differences?

> I would run an linear mixed effects model with age, condition, and their interaction predicting looking time. 


# Part 2: Simulation

```{r, warning=F, message=F}
library(tidyverse)
```

Let's start by convincing ourselves that t-tests have the appropriate false positive rate. Run 10,000 t-tests with standard, normally-distributed data from a made up 30-person, single-measurement experiment (the command for sampling from a normal distribution is `rnorm`).

The goal of these t-tests are to determine, based on 30 observations, whether the underlying distribution (in this case a normal distribution with mean 0 and standard deviation 1) has a mean that is different from 0. In reality, the mean is not different from 0 (we sampled it using `rnorm`), but sometimes the 30 observations we get in our experiment will suggest that the mean is higher or lower. In this case, we'll get a "significant" result and incorrectly reject the null hypothesis of mean 0.

What's the proportion of "significant" results ($p < .05$) that you see?

First do this using a `for` loop.

```{r}
# Parameters for the simulation
n_simulations <- 10000
sample_size <- 30
alpha <- 0.05
significant_results <- 0

for (i in 1:n_simulations) {
  # Sample from a normal distribution with mean = 0 and sd = 1
  sample_data <- rnorm(sample_size, mean = 0, sd = 1)
  # Perform a t-test against the null hypothesis (mean = 0)
  test_result <- t.test(sample_data, mu = 0)
  # If the p-value is less than alpha, increment the counter
  if (test_result$p.value < alpha) {
    significant_results <- significant_results + 1
  }
}

# Calculate the proportion of significant results
proportion_significant <- significant_results / n_simulations
proportion_significant

```

Next, do this using the `replicate` function:

```{r}
# Parameters for the simulation
n_simulations <- 10000
sample_size <- 30
alpha <- 0.05

# Use replicate to run the t-tests
set.seed(123) # Setting a seed for reproducibility
p_values <- replicate(n_simulations, {
  sample_data <- rnorm(sample_size, mean = 0, sd = 1)
  test_result <- t.test(sample_data, mu = 0)
  test_result$p.value
})

# Calculate the proportion of significant results
proportion_significant <- mean(p_values < alpha)
proportion_significant

```

How does this compare to the intended false-positive rate of $\alpha=0.05$?

> It's very close to it.

Ok, that was a bit boring. Let's try something more interesting - let's implement a p-value sniffing simulation, in the style of Simons, Nelson, & Simonsohn (2011).

Consider this scenario: you have done an experiment, again with 30 participants (one observation each, just for simplicity). The question is whether the true mean is different from 0. You aren't going to check the p-value every trial, but let's say you run 30 - then if the p-value is within the range p < .25 and p > .05, you optionally run 30 more and add those data, then test again. But if the original p value is < .05, you call it a day, and if the original is > .25, you also stop.  

First, write a function that implements this sampling regime.

```{r}
double.sample <- function() {
  # Sample the first batch of 30 observations
  data <- rnorm(30, mean = 0, sd = 1)
  # Perform the initial t-test
  initial_test <- t.test(data)
  
  # Check the p-value and decide whether to continue
  if (initial_test$p.value < 0.05 || initial_test$p.value > 0.25) {
    # If the p-value is < .05 or > .25, stop and return the result
    return(initial_test$p.value)
  } else {
    # If the p-value is between .05 and .25, sample another 30 observations
    data2 <- rnorm(30, mean = 0, sd = 1)
    # Combine the data and perform the final t-test
    final_test <- t.test(c(data, data2))
    return(final_test$p.value)
  }
}
```

Now call this function 10k times and find out what happens. 

```{r}
# Set up a variable to store the results
p.values <- replicate(10000, double.sample())

# Calculate the proportion of significant results (p < .05)
proportion_significant <- mean(p.values < 0.05)

# Print the result
proportion_significant
```

Is there an inflation of false positives? How bad is it?

> Yes! 0.073/0.0465 ~= 1.5 times more! 

Now modify this code so that you can investigate this "double the sample" rule in a bit more depth. In the previous question, the researcher doubles the sample only when they think they got "close" to a significant result, i.e. when their not-significant p is less than 0.25. What if the researcher was more optimistic? See what happens in these 3 other scenarios:

* The researcher doubles the sample whenever their pvalue is not significant, but it's less than 0.5.
* The researcher doubles the sample whenever their pvalue is not significant, but it's less than 0.75.
* The research doubles their sample whenever they get ANY pvalue that is not significant.

How do these choices affect the false positive rate?

HINT: Try to do this by making the function `double.sample` take the upper p value as an argument, so that you can pass this through dplyr.

HINT 2: You may need more samples. Find out by looking at how the results change from run to run.

```{r}

double.sample <- function(p.upper.limit) {
  # Sample the first batch of 30 observations
  data <- rnorm(30, mean = 0, sd = 1)
  # Perform the initial t-test
  initial_test <- t.test(data)
  
  # Check the p-value against the upper limit
  if (initial_test$p.value < 0.05 || initial_test$p.value > p.upper.limit) {
    # If the p-value is < .05 or > p.upper.limit, stop and return the result
    return(initial_test$p.value)
  } else {
    # If the p-value is between .05 and p.upper.limit, sample another 30 observations
    data2 <- rnorm(30, mean = 0, sd = 1)
    # Combine the data and perform the final t-test
    final_test <- t.test(c(data, data2))
    return(final_test$p.value)
  }
}

# Run the simulation with different p-value upper limits
set.seed(123) # for reproducibility
p.upper.limits <- c(0.25, 0.5, 0.75, 1)
results <- tibble(p.upper.limit = p.upper.limits) %>%
  mutate(
    p.values = map(p.upper.limit, ~replicate(100000, double.sample(.x))),
    proportion_significant = map_dbl(p.values, ~mean(.x < 0.05))
  )

# Check the results
results

```

What do you conclude on the basis of this simulation? How bad is this kind of data-dependent policy?

> These policies inflate false positives! Specifically, if you decrease your threshold to double the sample and rerun results, you'll get more false positives! 

